RegressionPvNP

### 2017-08-02 RegressionPvNP.Rmd
Wrapping up model

includeDemo = FALSE # demographics

@@

cache.dir = paste(data, 'Analysis',sub('\\+', '', tile.w), sep='/')
if (!dir.exists(cache.dir)) dir.create(cache.dir)
```

@@

unit.var <- function(x) {
  if (sd(x) == 0) return(x)
  return((x-mean(x))/sd(x) )
}

#dysplasia.df = apply(dysplasia.df, 2, function(x) x-median(x))
dysplasia.df = apply(dysplasia.df, 2, unit.var)

@@

  ## Some sort of complexity score
  #plot(unlist(complexity.measures), col=sum.patient.data$Status)
  # t.test(
  # complexity.measures[subset(sum.patient.data, Status == 'P')$Hospital.Research.ID],
  # complexity.measures[subset(sum.patient.data, Status == 'NP')$Hospital.Research.ID]
  # )

  score.cx <-function(pd) {
    samples = grep('^D', colnames(pd$seg.vals), value=T)
    len.var = var(pd$seg.vals$end.pos - pd$seg.vals$start.pos) # length won't vary between samples

    #extremes = apply(as.matrix(pd$seg.vals[,samples]), 2, function(col) {
    #  length(which(col >= (median(col)+2*sd(col)))) + length(which(col <= (median(col)-2*sd(col)))) 
    #})

    num.var.reg = vector(mode='integer', length=length(samples))
    if (!is.null(pd$variable.regions) && nrow(pd$variable.regions) > 0) {
      num.var.reg = apply(as.matrix(pd$variable.regions[,samples]), 2, function(col) {
        length(which(col >= 1.1 | col <= 0.9))
      })
    }

    sample.var = apply(as.matrix(pd$seg.vals[, samples]), 2, var)
    t(cbind(len.var, num.var.reg, sample.var))
  }

  complexity.measures = lapply(patient.data, score.cx)

  dysplasia.df = cbind(dysplasia.df, 'segment.length'=0, 'num.var.regions'=0, 'sample.variance'=0)
  for (pt in names(complexity.measures)) {
    for (sample in colnames(complexity.measures[[pt]])) {
      dysplasia.df[sample,c('segment.length', 'num.var.regions', 'sample.variance')] = as.numeric(complexity.measures[[pt]][,sample])
    }
    #for (sample in rownames(dysplasia.df)) {
      #dysplasia.df[sample, ] = 
        #complexity.measures[subset(patient.info, Samplename == sample)$Hospital.Research.ID]
    #}
  }
  # norm to 1 to keep within the magnitude of the genomic data
  scale.to.1<-function(x) {
    (x-min(x))/(max(x)-min(x))
  }

  dysplasia.df[,c('segment.length', 'num.var.regions', 'sample.variance')] = apply(dysplasia.df[,c('segment.length', 'num.var.regions', 'sample.variance')], 2, scale)

  #cx = dysplasia.df[,'cx.measure'] 
  #dysplasia.df[,'cx.measure'] = (log(cx)/log(max(cx)))
  #dysplasia.df[,'cx.measure'] = dysplasia.df[,'cx.measure'] - median(dysplasia.df[,'cx.measure'])
  #dysplasia.df[,'cx.measure'] = ((cx - min(cx))/(max(cx)-min(cx)))

@@

  # Encode based on presumed distributions or unit normalize
  demo.data = demo.data %>% rowwise() %>% mutate( 'age.encoded'= encode.age(Age.at.diagnosis))
  demo.data$Age.at.diagnosis = demo.data$age.encoded
  demo.data$age.encoded = NULL
  demo.data$BMI = scale(demo.data$BMI)

  demo.data[,c('Circumference','Maximal')] = apply(demo.data[,c('Circumference','Maximal')], 2, function(x) (x-mean(x, na.rm=T))/sd(x, na.rm=T))

@@

    cv.class = matrix(nrow=splits, ncol=length(lambda))
    deviance = matrix(nrow=splits, ncol=length(lambda))
    for (i in 1:splits) { 
      message(paste(i, "split"))
      test.rows = which(rownames(x) %in% tpts[which(tpts[,setCol] == i), 'Samplename'])
      test = x[test.rows,]
      training = x[-test.rows,]
      # pre-spec lambda seq
      fitCV = glmnet(training, y[-test.rows], lambda=lambda, alpha=a, family='binomial', ...) 
      # autoplot(fit) + theme(legend.position="none")

@@

    lmin = subset(df, lambda.min)$lambda
    df[subset(df, log.lambda <= log(lmin) + log(sd(lambda))/2 )[1, 'lambda.at'], 'lambda.1se'] = T
  }

@@

```{r xvalpt, message=F, warning=F, echo=F, fig.height=16, fig.width=16}
# dysplasia.df = cbind(dysplasia.df, 'sl.nvr'=0)
# dysplasia.df[, 'sl.nvr'] = scale(dysplasia.df[, "segment.length"] + dysplasia.df[, 'num.var.regions'])
# 
# dysplasia.df = cbind(dysplasia.df, 'sl.nvr.sv'=0)
# dysplasia.df[, 'sl.nvr.sv'] = scale(dysplasia.df[, "segment.length"] + dysplasia.df[, 'num.var.regions'] + dysplasia.df[, 'sample.variance'])
# 
# dysplasia.df = cbind(dysplasia.df, 'sl.sv'=0)
# dysplasia.df[, 'sl.sv'] = scale(dysplasia.df[, "segment.length"] + dysplasia.df[, 'sample.variance'])
# 
# dysplasia.df = cbind(dysplasia.df, 'nvr.sv'=0)
# dysplasia.df[, 'sl.sv'] = scale(dysplasia.df[, "num.var.regions"] + dysplasia.df[, 'sample.variance'])

nl = 1000

@@

file = paste(cache.dir, 'patient_folds.tsv', sep='/')
if (file.exists(file)) {
  sets = read.table(file, header=T, sep='\t')
} else {
  sets = create.patient.sets(pts, folds, splits, 0.2)  
  write.table(sets, quote=F, sep='\t', row.names=F, file=file)
}

@@

    l = fit0$lambda
    if (a == 0) {
      l = sort(c(fit0$lambda, seq(exp(-5), exp(-10), -1e-6),
            seq(exp(-10), exp(-15), -1e-8),
            seq(exp(-15), exp(-20), -1e-9)), decreasing=T)
    }
    cv.patient = crossvalidate.by.patient(x=dysplasia.df, y=labels, lambda=l, pts=sets, a=a, nfolds=folds, splits=splits, fit=fit0, select='deviance', standardize=F)

@@

```{r echo=F, warning=F, message=F, eval=F}
### Features in samples
## TODO Look at the coefficients, are the high pos ones "gains" and the low neg ones "losses" etc
plot(coefs[[select.alpha]][,1])
selected = coefs[[select.alpha]]

selected[which(
  selected[,1] > mean(selected[,1]) + sd(selected[,1]) | 
  selected[,1] < mean(selected[,1] - sd(selected[,1]))  
),]

## Variance in the selected features?  Not so much
coefs[[select.alpha]][,1]

v = vector(length(colnames(dysplasia.df)), mode='numeric')
names(v) = colnames(dysplasia.df)
v[rownames(selected)] = coefs[[select.alpha]][,1]

vv <- apply(dysplasia.df, 2, var)
plot(v, vv, log='y', col=(v!=0)+1)

boxplot(vv ~ (v==0), log='y', ylim=c(0.0001,1))

qqplot(vv[v==0], vv[v!=0], log='xy')

length(which( coef.stable[[select.alpha]] >= 0.5) )

regions = rownames(coefs[[select.alpha]])

pg = dysplasia.df[ labels == 1 ,regions]

nonpg = dysplasia.df[ labels == 0 ,regions]

colnames(pg)

meantests = sapply(regions, function(r)   t.test(pg[,r], nonpg[,r])$p.value )

which(meantests < 0.05)

#summary(pg[,'1:119640299-129610322'])
#summary(nonpg[,'1:119640299-129610322'])

@@

```{r, warning=F, message=F, eval=T}
#topF = names(which((coef.stable[['1']]) >= 0.75))
topF = names(coef.stable[[select.alpha]])

@@

```{r, echo=F, warning=F, message=F}
chromInfo = read.table('hg19_genes.txt', sep='\t', header=T)
chromInfo$chr = factor(chromInfo$chr, levels=c(1:22, 'X','Y'))

@@

```{r, warning=F, message=F}
bm = NULL
tryCatch({
  mart = biomaRt::useMart("ENSEMBL_MART_ENSEMBL", host="grch37.ensembl.org")
  biomart = useDataset("hsapiens_gene_ensembl", mart)

@@

```{r echo=F, warning=F, message=F,eval=T, fig.height=8, fig.width=8}

hazards = apply( exp(t(dysplasia.df[, rownames(coefs[[select.alpha]])]) *  coefs[[select.alpha]][,1]), 1, function(x) {
  sd(x)/mean(x)
})
ch = as.data.frame(hazards)
ch$coef = coefs[[select.alpha]][,1]
ch$label = rownames(ch)
ch$stability = featureStability

ggplot(ch, aes(coef, hazards, col=hazards > quantile(hazards)[['75%']])) + 
  #xlim(min(coefGR$coef)-5, max(coefGR$coef)+10) +
  geom_point(alpha=0.5) + 
  geom_text_repel(aes(x=coef, label=ifelse(hazards > quantile(hazards)[['75%']], label, '')),show.legend=F) +
  labs(title='Top features coef vs hazard') + theme(legend.position = 'bottom')
```

```{r echo=F, warning=F, message=F,eval=F, fig.height=12, fig.width=8}
ch = arrange(ch, -stability, -hazards, -coef)
ch$hazards = round(ch$hazards, 3)
ch$coef = round(ch$coef, 3)
write.table(ch[c('label', 'stability','hazards','coef')], row.names = F, quote=F, sep="\t", file='features.txt')

  reg = cbind.data.frame( 'value'=dysplasia.df[ ,r], labels)
  reg$labels = factor(reg$labels)
  reg$labels = revalue(reg$labels, c("1"="P", "0"="NP"))

plot.reg<-function(x) {
  reg = melt.by.reg(x)

  pv = ifelse(reg$pval[1] < .001, '< .001', format.pval(unique(reg$pval[1])))

#  ggplot(reg, aes(value, fill=labels) ) + geom_histogram(bins=50, color='grey') + 
#    labs(x='seg. value', title=r, subtitle=paste('haz=',x[['hazards']], ' coef=', x[['coef']], ' MW pval: ',pv,sep=''))
  ggplot(reg, aes(x=labels, y=value, group=labels, fill=labels) ) + geom_jitter(color='grey', show.legend = F) + geom_violin(show.legend = F, alpha=0.5) +
    geom_hline(yintercept = median(reg$value) + sd(reg$value)) + geom_hline(yintercept = median(reg$value) - sd(reg$value)) +
    labs(x='', y='seg. value', title=x[['label']], subtitle=paste('haz= ',x[['hazards']], ' coef= ',x[['coef']], '\nMW pval ',pv,sep=''))
}

do.call(grid.arrange, c(lapply(regions[1:15], function(reg) {
  print(reg$region[1])
  cols = c('indianred3','slateblue2')
  if ( as.numeric(reg$coef[1]) < 0) cols = c('green3', 'darkviolet')

  ggplot(reg, aes(labels, value*coef, fill=labels, color=labels)) + 
    geom_jitter(aes(labels, value), width=0.05, color='yellow4', alpha=0.3) +
    geom_jitter(aes(labels, value*coef, color=labels), width=0.05, alpha=0.5) +
    geom_text_repel(data=subset(reg, value >= quantile(reg$value)[['100%']]), aes(labels, value, label=round(value, 3)), color='yellow4') +
    geom_text_repel(data=subset(reg, value >= quantile(reg$value)[['100%']]), aes(labels, value*coef, label=round(value, 3),color=labels)) + geom_violin(alpha=0.2) + 
    scale_fill_manual(values=cols) + scale_color_manual(values=cols) +
    theme(legend.position = 'none') + labs(title=unique(reg$region)[1], x='')
})))

    #geom_violin(aes(labels, value, fill=labels), alpha=0.5) + theme(legend.position = 'none')

#  x = rbind.data.frame('P'= cbind( 'gain' = length(which(with(subset(reg, labels == 'P'), value > 0))),
#                               'loss' = length(which(with(subset(reg, labels == 'P'), value < 0))) ),

#                        'NP' = cbind( 'gain' = length(which(with(subset(reg, labels == 'NP'), value > 0))),
#                                'loss' = length(which(with(subset(reg, labels == 'NP'), value < 0))) ) )

    # x = rbind.data.frame('P'= cbind( 'gain' = length(which(with(subset(reg, labels == 'P'), value >= median(value) + sd(value)*se))),
  #                              'loss' = length(which(with(subset(reg, labels == 'P'), value <= median(value) - sd(value)*se))) ),
  # 
  #                 'NP' = cbind( 'gain' = length(which(with(subset(reg, labels == 'NP'), value >= median(value) + sd(value)*se))),
  #                               'loss' = length(which(with(subset(reg, labels == 'NP'), value <= median(value) - sd(value)*se))) ) )
  # x = rbind.data.frame('P'= cbind( 'high' = length(which(with(subset(reg, labels == 'P'), value >= 1.1))),
  #                                  'low' = length(which(with(subset(reg, labels == 'P'), value <= 0.9))) ),
  # 
  #                 'NP'= cbind( 'high' = length(which(with(subset(reg, labels == 'NP'), value >= 1.1))),
  #                                  'low' = length(which(with(subset(reg, labels == 'NP'), value <= 0.9))) ))

  #cols = c('indianred3','slateblue2')
  #if ( as.numeric(reg$coef[1]) < 0) cols = c('plum', 'skyblue3')

  #ggplot(melt(as.matrix(x)), aes(Var2, value, fill=Var1)) + facet_grid(~Var1) +
  #  geom_col( show.legend = F) + labs(x='', y='', title=unique(reg$region)[1]) + scale_fill_manual(values=cols)

#})))

## Individual hazards vs CN values?  exp(coef) -- odds ratio
# plot( exp( coefs[[select.alpha]][ch$label[1],1]  * dysplasia.df[,ch$label[1]]),  dysplasia.df[,ch$label[1]] )
# 
# plot(  exp( t(dysplasia.df[, ch$label[1]]) *  coefs[[select.alpha]][ch$label[1],1]  ), dysplasia.df[,ch$label[1]]  ) ## pos coef
# plot(exp(t(dysplasia.df[, ch$label[3]]) *  coefs[[select.alpha]][ch$label[3],1]), dysplasia.df[,ch$label[3]]) ## pos coef
# 
# plot(exp(t(dysplasia.df[, ch$label[2]]) *  coefs[[select.alpha]][ch$label[2],1]), dysplasia.df[,ch$label[2]]) ## neg coef
# plot(exp(t(dysplasia.df[, ch$label[4]]) *  coefs[[select.alpha]][ch$label[4],1]), dysplasia.df[,ch$label[4]]) ## neg coef
# 
# 
# labels[which( dysplasia.df[, ch$label[2]] <= 0.9)]
# 
# labels[which( dysplasia.df[, ch$label[1]] >= 1.1)]

#do.call(grid.arrange, c(apply(ch[1:10,],1, plot.reg), ncol=2))

#do.call(grid.arrange, c(apply(ch[11:20,],1, plot.reg), ncol=2))

#do.call(grid.arrange, c(apply(ch[21:30,],1, plot.reg), ncol=2))

```{r}

cols = intersect(rownames(ch), grep('\\d+:', colnames(dysplasia.df), invert=T, value=T))
if (length(cols) > 0)
  pander(ch[cols,1:2], caption='Non-zero demographic hazard ratios')
```

`r pander(ch[order(-ch$hazards),][1:10,1:2], caption='Top 10 features by hazard ratio.')`

```{r echo=F, warning=F, message=F}
if (!is.null(bm)) {
  topFRegions = merge(topFRegions, (bm %>% group_by(chromosome_name) %>% summarise('gene.cnt'=length(unique(hgnc_symbol)), 'exon.cnt'=sum(exon_count))), by.x='chr', by.y='chromosome_name', all.x=T)
  colnames(topFRegions) = sub('\\.(x|y)', '', colnames(topFRegions)  )

@@

@@

@@

  write.table(unlist(strsplit(unique(coefGR$COSMIC_ensembl_ids), ',')), quote=F, row.names = F, col.names = F, file='feature_cosmic.txt')
}
#However, the `r round(bpCovered, 2)*100`% of bases appear to include `r round(length(unique(bm$hgnc_symbol))/sum(chromInfo$protein_coding), 2)*100`% (`r length(unique(bm$hgnc_symbol))`) of known protein-coding genes.  Similarly, `r round(length(unlist(strsplit(unique(coefGR$COSMIC), ',')))/length(ccgenes$Gene.Symbol),2)*100`% of the `r length(ccgenes$Gene.Symbol)` COSMIC genes are contained within these regions.

@@

try

@@

  save(gp, lp, both, file=file)
}

do.call(grid.arrange, c(gp$plots, top='All samples, GAINS 10fold, 5 splits'))
do.call(grid.arrange, c(lp$plots, top='All samples, LOSSES 10fold, 5 splits'))
do.call(grid.arrange, c(both$plots, top='All samples, gains & losses 10fold, 5 splits'))

@@

  # sort in label order
  r.dysplasia.df = t(mergedDf[,names(r.labels)])

@@

    train.rows = which(rownames(dysplasia.df) %in% samples)
    training = dysplasia.df[train.rows,]
    test = as.matrix(dysplasia.df[-train.rows,])
    if ( nrow(test) == ncol(dysplasia.df) ) test = t(test)

@@

@@

```{r validation, message=F, warning=F, echo=F, fig.height=8, fig.width=6}

score.cx <-function(pd) { ## OLD method
#  print(pd$info)
  len.var = (var(pd$seg.vals$end.pos - pd$seg.vals$start.pos))/1e12
  extreme.values = max( apply(as.matrix(pd$seg.vals[,grep('^D', colnames(pd$seg.vals))]), 2, function(col) {
    length(which(col >= (median(col)+2*sd(col)))) + length(which(col <= (median(col)-2*sd(col)))) 
  }) )

  (len.var + extreme.values)
}

@@

complexity.measures = sapply(validation.patient.data, score.cx)
validation.df = cbind(validation.df, 'cx.measure'=NA)
for (sample in rownames(validation.df)) {
  validation.df[sample, 'cx.measure'] = complexity.measures[subset(validation.patient.info, Samplename == sample)$Hospital.Research.ID]
}
# 
# # norm to 1 to keep within the magnitude of the genomic data
cx = validation.df[,'cx.measure']
validation.df[,'cx.measure'] = (log(cx)/log(max(cx)))

@@

#fitV <- glmnet(dysplasia.df, labels, alpha=select.alpha, nlambda=nl, family='binomial') # all patients
#autoplot(fitV)
fitV = models[[select.alpha]]
lambda.opt = performance[select.alpha, 'lambda']

@@

```{r, echo=F, message=F, warning=F, fig.width=8, fig.height=12}

@@

grid.arrange(p1, p2,ncol=1)
```

```{r echo=F, warning=F, message=F}

>>>>

includeDemo = F # demographics

@@

cache.dir = paste(data, 'Analysis',sub('\\+', '', tile.w),   sep='/')

if (!dir.exists(cache.dir)) dir.create(cache.dir)
```

@@

unit.var <- function(x) {
  if (sd(x, na.rm=T) == 0) return(x)
  return((x-mean(x,na.rm=T))/sd(x,na.rm=T) )
}

dysplasia.df = apply(dysplasia.df, 2, unit.var)

@@

  score.cx <- function(pt.d, df) {
    get.length.variance <- function(pd) {
      samples = grep('^D', colnames(pd$seg.vals), value=T)
      len.var = var(pd$seg.vals$end.pos - pd$seg.vals$start.pos) # length won't vary between samples

      sample.var = apply(as.matrix(pd$seg.vals[, samples]), 2, var)
      t(cbind(len.var, sample.var))
    }

    complexity.measures = lapply(pt.d, get.length.variance)

    df = cbind(df, 'cx' = apply(df, 1, function(x) length(which(x >= sd(x)*2 | x <= -sd(x)*2))))
    df[,'cx'] = unit.var(df[,'cx'])

    df = cbind(df, 'segment.length'=0, 'sample.variance'=0)
    for (pt in names(complexity.measures)) {
      for (sample in colnames(complexity.measures[[pt]])) {
        df[sample,c('segment.length', 'sample.variance')] = as.numeric(complexity.measures[[pt]][,sample])
      }
    }

   df[, 'segment.length'] = unit.var(df[, 'segment.length'])
   df[, 'sample.variance'] = unit.var(df[, 'sample.variance'])

  return(df)
  }

## TODO score complexity on the seg.vals rather than merged data frame
  dysplasia.df = score.cx(patient.data, dysplasia.df)

@@

  # Encode based on presumed distributions or unit normalize
  demo.data = demo.data %>% rowwise() %>% mutate( 'age.encoded'= encode.age(Age.at.diagnosis))
  demo.data$Age.at.diagnosis = demo.data$age.encoded
  demo.data$age.encoded = NULL
  demo.data$BMI = unit.var(demo.data$BMI)

  demo.data[,c('Circumference','Maximal')] = apply(demo.data[,c('Circumference','Maximal')], 2, unit.var)

@@

    cv.class = matrix(nrow=splits, ncol=length(lambda))
    deviance = matrix(nrow=splits, ncol=length(lambda))
    for (i in 1:splits) { 
      message(paste(i, "split"))
      test.rows = which(rownames(x) %in% tpts[which(tpts[,setCol] == i), 'Samplename'])
      test = x[test.rows,]
      training = x[-test.rows,]
      # pre-spec lambda seq
      fitCV = glmnet(training, y[-test.rows], lambda=lambda, alpha=a, family='binomial', standardize = F) 
      # autoplot(fit) + theme(legend.position="none")

@@

    lmin = subset(df, lambda.min)$lambda
    lambdas = subset(df, log.lambda <= log(lmin) + log(sd(lambda))/2 )
    n = 3
    while (nrow(lambdas) <= 0 & n < 10) {
      lambdas = subset(df, log.lambda <= log(lmin) + log(sd(lambda))/n )
      n = n+1
    }
    if (nrow(lambdas) <= 0) {
      df[subset(df, log.lambda <= log(lmin) )[1, 'lambda.at'], 'lambda.1se'] = T  
    } else {
      df[subset(df, log.lambda <= log(lmin) + log(sd(lambda))/n )[1, 'lambda.at'], 'lambda.1se'] = T  
    }
  }

@@

```{r xvalpt, message=F, warning=F, echo=F, fig.height=16, fig.width=16}
nl = 1000

@@

file = paste(data, 'Analysis/patient_folds.tsv', sep='/')
if (file.exists(file)) {
  message(paste("Reading folds file", file))
  sets = read.table(file, header=T, sep='\t')
} else {
  #sets = create.patient.sets(pts, folds, splits, 0.2)  
  #write.table(sets, quote=F, sep='\t', row.names=F, file=file)
  stop("Missing patient folds file")
}

@@

    l = fit0$lambda
    if (a == 0) {
      l = sort(c(fit0$lambda, seq(exp(-5), exp(-10), -1e-6),
            seq(exp(-10), exp(-15), -1e-8),
            seq(exp(-15), exp(-20), -1e-9)), decreasing=T)
    }
    cv.patient = crossvalidate.by.patient(x=dysplasia.df, y=labels, lambda=l, pts=sets, a=a, nfolds=folds, splits=splits, fit=fit0, select='deviance', standardize=F)
    #cv.patient$deviance.plot

@@

@@

```{r featureinfo, warning=F, message=F, eval=T}
#topF = names(which((coef.stable[['1']]) >= 0.75))
topF = names(coef.stable[[select.alpha]])

@@

```{r bplen, echo=F, warning=F, message=F}
chromInfo = read.table('hg19_genes.txt', sep='\t', header=T)
chromInfo$chr = factor(chromInfo$chr, levels=c(1:22, 'X','Y'))

@@

```{r genes, warning=F, message=F}
bm = NULL
tryCatch({
  mart = biomaRt::useMart("ENSEMBL_MART_ENSEMBL", host="grch37.ensembl.org")
  biomart = useDataset("hsapiens_gene_ensembl", mart)

@@

```{r hazards, echo=F, warning=F, message=F,eval=T, fig.height=12, fig.width=12}

hazards = apply( exp(t(dysplasia.df[, rownames(coefs[[select.alpha]])]) *  coefs[[select.alpha]][,1]), 1, function(x) {
  sd(x)/mean(x)
})
ch = as.data.frame(hazards)
ch$coef = coefs[[select.alpha]][,1]
ch$stability = featureStability
ch$label = rownames(ch)

qq = quantile(ch$hazards, seq(0,1,.15))
ggplot(ch, aes(coef, hazards, col=hazards >= qq[length(qq)])) + 
  geom_point(alpha=0.5) + 
  geom_text_repel(aes(x=coef, label=ifelse(hazards >= qq[length(qq)], label, '')),show.legend=F) +
  labs(title='Top features by hazard ratio') + theme(legend.position = 'bottom') + theme(legend.position='none')
```

```{r, echo=F}
cols = intersect(rownames(ch), grep('\\d+:', colnames(dysplasia.df), invert=T, value=T))
if (length(cols) > 0)
pander(ch[cols,1:2], caption='Non-zero demographic hazard ratios')
```

#### COSMIC genes

```{r cosmic, echo=F, warning=F, message=F}
if (!is.null(bm)) {
  topFRegions = merge(topFRegions, (bm %>% group_by(chromosome_name) %>% summarise('gene.cnt'=length(unique(hgnc_symbol)), 'exon.cnt'=sum(exon_count))), by.x='chr', by.y='chromosome_name', all.x=T)
  colnames(topFRegions) = sub('\\.(x|y)', '', colnames(topFRegions)  )

@@

  coefGR$stability = rowSums(coefs[[select.alpha]][grcfs,-1])/(folds*splits)

@@

write.table(as.data.frame(coefGR[order(-coefGR$stability)])[, c('seqnames', 'start','end','coef','stability','hazard', 'COSMIC')], sep='\t', quote=F, row.names=F, file='cosmic_genes.txt')

@@

write.table(unlist(strsplit(unique(coefGR$COSMIC_ensembl_ids), ',')), quote=F, row.names = F, col.names = F, file='feature_cosmic.txt')
}
#However, the `r round(bpCovered, 2)*100`% of bases appear to include `r round(length(unique(bm$hgnc_symbol))/sum(chromInfo$protein_coding), 2)*100`% (`r length(unique(bm$hgnc_symbol))`) of known protein-coding genes.  Similarly, `r round(length(unlist(strsplit(unique(coefGR$COSMIC), ',')))/length(ccgenes$Gene.Symbol),2)*100`% of the `r length(ccgenes$Gene.Symbol)` COSMIC genes are contained within these regions.
```

```{r, echo=F}
pander(as.data.frame(coefGR[coefGR$COSMIC != ""])[c(1:3, 11)], caption='COSMIC genes mapped to coefficients. Only a few regions have mapped COSMIC genes, but several regions neighbor those with important tumour supressors (TP53) and may have not been selected through an artifact of the GLM process.')
```

### How do coefficients related to the raw data?

```{r warning=F, message=F, fig.height=8, fig.width=8}
ch = arrange(ch, -stability, -hazards, -coef)
ch$hazards = round(ch$hazards, 3)
ch$coef = round(ch$coef, 3)
write.table(ch[c('label', 'stability','hazards','coef')], row.names = F, quote=F, sep="\t", file='features.txt')

  reg = cbind.data.frame( 'value'=dysplasia.df[ ,r], labels)
  reg$labels = factor(reg$labels, levels=c(1,0), ordered=T)
  reg$labels = revalue(reg$labels, c("1"="P", "0"="NP"))

  #reg = factor(reg$labels, levels=c('P','NP'), ordered=T)

# plot.reg<-function(x) {
#   reg = melt.by.reg(x)
#   
#   pv = ifelse(reg$pval[1] < .001, '< .001', format.pval(unique(reg$pval[1])))
# 
#   ggplot(reg, aes(x=labels, y=value, group=labels, color=labels, fill=labels) )  + geom_violin(alpha=0.5) + geom_jitter(width=0.3, alpha=0.8) +
#     geom_hline(yintercept=with(reg, c(mean(value)+sd(value)*2, mean(value)-sd(value)*2)), color='darkgrey', linetype='dashed') +
#     #geom_hline(yintercept = mean(reg$value) + sd(reg$value)) + geom_hline(yintercept = mean(reg$value) - sd(reg$value)) +
#     labs(x='', y='seg. value', title=x[['label']], subtitle=paste('haz= ',x[['hazards']], ' coef= ',x[['coef']], '\nMW pval ',pv,sep='')) + theme(legend.position = 'none')
# }

plot.features<-function(reg) {
  se = 2
  cols = c('indianred3','slateblue2')
  if ( as.numeric(reg$coef[1]) < 0) cols = c('green3', 'darkviolet')

  mn = 0
  if (min(reg$value) >= 0) {
    mn = mean(reg$value)  
  }

  p = ggplot(reg, aes(labels, value, group=labels, fill=labels, color=labels)) + geom_violin(alpha=0.2) +
    geom_jitter(width=0.05) +
    geom_hline(yintercept=with(reg, c(mn+sd(value)*se, mn-sd(value)*se)), color='darkgrey', linetype='dashed') +
    scale_fill_manual(values=cols) + scale_color_manual(values=cols) + theme(legend.position = 'none') + labs(x='', subtitle=reg$coef[1])

   x = rbind.data.frame('P'= cbind( 'gain' = length(which(with(subset(reg, labels == 'P'), value >=  mn+sd(value)*se))),
                                    'loss' = -length(which(with(subset(reg, labels == 'P'), value <=  mn-sd(value)*se))) ),

                        'NP' = cbind( 'gain' = length(which(with(subset(reg, labels == 'NP'), value >=  mn+sd(value)*se))),
                                      'loss' = -length(which(with(subset(reg, labels == 'NP'), value <=  mn-sd(value)*se))) ) )

   x = x*sign(reg$coef[1])

  p2 = ggplot(melt(as.matrix(x)), aes(Var2, value, fill=Var1, 
                                      color=ifelse(value > 0, 'red', 'white'))) + #facet_grid(~Var1) +
        geom_bar(stat='identity', width=0.5, position='dodge') + 
        geom_hline(yintercept = 0, color='black', size=1.5) +
        labs(x='', y='Counts', subtitle=reg$coef[1]) + 
        scale_fill_manual(values=cols) + theme(legend.position = 'none')

  return(list('plot'=arrangeGrob(p, p2, ncol=2, top=unique(reg$region)[1]), 'counts'=x))     
}

pf = lapply(regions, plot.features)

do.call(grid.arrange, c(lapply(pf[1:12], function(x) x$plot), ncol=3) )

### Is there a difference in P~NP for gains/losses

Based on the relationship between the coefficients and the values of the data we would expect one.  
```{r}
# positive coefficients were related to gains in progressors
gains = subset(ch, coef > 0)
# negative coefficients were related to losses in non-progressors
losses = subset(ch, coef < 0)

gain = do.call(rbind, lapply(gains$label, function(reg) {
  values = dysplasia.df[,reg] 
  table(labels[which(values > sd(values))])
}))

# There is a difference in the gains NP~P in the positive coefficients
wilcox.test(gain[,1], gain[,2])
summary(aov(gain[,1]~gain[,2]))

loss = do.call(rbind, lapply(gains$label, function(reg) {
  values = dysplasia.df[,reg] 
  table(labels[which(values < -sd(values))])
}))

# There may be on in the losses as well, but less clear
wilcox.test(loss[,1], loss[,2])
summary(aov(loss[,1]~loss[,2]))

# Compared to randomly selected regions (not the coefficients)
sample.reg = sample(setdiff(colnames(dysplasia.df), ch$label), nrow(ch))

randomGains = do.call(rbind, lapply(sample.reg, function(reg) {
  values = dysplasia.df[,reg] 
  table(labels[which(values > sd(values))])
}))

# P~NP may be different across randomly selected regions as well, again less clear
wilcox.test(randomGains[,1], randomGains[,2])
summary(aov(randomGains[,1]~randomGains[,2]))

randomLosses = do.call(rbind, lapply(sample.reg, function(reg) {
  values = dysplasia.df[,reg] 
  table(labels[which(values < -sd(values))])
}))

# In this case losses show a definite difference
wilcox.test(randomLosses[,1], randomLosses[,2])
summary(aov(randomLosses[,1]~randomLosses[,2]))

@@

If gains or losses are primarily driving this, can a model that only looks at gains or losses predict better?

@@

  save(gp, lp, both, file=file)
}
```

Gains alone are pretty much the same as the raw data.
```{r warning=F, message=F, fig.height=8, fig.width=8}
do.call(grid.arrange, c(gp$plots, top='All samples, GAINS 10fold, 5 splits'))
```

So are losses...
```{r warning=F, message=F, fig.height=8, fig.width=8}
do.call(grid.arrange, c(lp$plots, top='All samples, LOSSES 10fold, 5 splits'))
```

When we try to put them together, so each region has two columns where gains and losses are considered separately the classification drops to < 50%. So we're better off using the merged data we started with and gains/losses alone aren't simply driving the prediction.
```{r warning=F, message=F, fig.height=8, fig.width=8}
do.call(grid.arrange, c(both$plots, top='All samples, gains & losses 10fold, 5 splits'))
```

@@

  # sort in label order
  r.dysplasia.df = dysplasia.df[names(r.labels),]

@@

    train.rows = which(rownames(dysplasia.df) %in% samples)
    training = dysplasia.df[train.rows,]
    test = as.matrix(dysplasia.df[-train.rows,])
    if (ncol(test) <= 1) next
    #if ( nrow(test) == ncol(dysplasia.df) ) test = t(test)

@@

by.path = do.call(rbind.data.frame, lapply(pg.samp[subset(sum.patient.data, Status == 'NP')$Hospital.Research.ID], function(pt) {
  final.endo = subset(pt, Endoscopy.Year == max(Endoscopy.Year))
  final.endo = final.endo[nrow(final.endo),]

  pred <- function(rows) length(which(rows$Prediction > cutoff))
  if (pt$Status[1] == 'NP')
    pred <- function(rows) length(which(rows$Prediction < cutoff))

  cbind('HGD'=pred(h), 'total.h'=nrow(h), 
        'LGD'=pred(l), 'total.l'=nrow(l), 
        'ID'=pred(id), 'total.id'=nrow(id),
        'NDBE'=pred(be), 'total.be'=nrow(be))
}))

sum(by.path$NDBE)/sum(by.path$total.be)
sum(by.path$LGD)/sum(by.path$total.l)
sum(by.path$ID)/sum(by.path$total.id)
sum(by.path$HGD)/sum(by.path$total.h)

@@

```{r validation, message=F, warning=F, echo=F, fig.height=8, fig.width=6, eval=F}

@@

validation.df = score.cx(validation.patient.data, validation.df)

@@

#fitV <- glmnet(dysplasia.df, labels, alpha=select.alpha, nlambda=nl, family='binomial') # all patients
#autoplot(fitV)
fitV = models[[select.alpha]]
lambda.opt = performance.at.1se[[select.alpha]][, 'lambda']

@@

```{r, echo=F, message=F, warning=F, fig.width=8, fig.height=12, eval=F}

@@

grid.arrange(p1, p2,ncol=1)

by.path = do.call(rbind.data.frame, lapply(vpd, function(pt) {
  final.endo = subset(pt, Endoscopy.Year == max(Endoscopy.Year))
  final.endo = final.endo[nrow(final.endo),]

  pred <- function(rows) length(which(rows$Prediction > cutoff))
  if (pt$Status[1] == 'NP')
    pred <- function(rows) length(which(rows$Prediction < cutoff))

  cbind('HGD'=pred(h), 'total.h'=nrow(h), 
        'LGD'=pred(l), 'total.l'=nrow(l), 
        'ID'=pred(id), 'total.id'=nrow(id),
        'NDBE'=pred(be), 'total.be'=nrow(be))
}))

sum(by.path$HGD)/sum(by.path$total.h)
sum(by.path$LGD)/sum(by.path$total.l)
sum(by.path$ID)/sum(by.path$total.id)
sum(by.path$NDBE)/sum(by.path$total.be)

```{r echo=F, warning=F, message=F, eval=F}
